Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to agroecossystems

### 2021-08-05 Manuscript/Manuscript.Rmd
JH

In the southeastern US, early growing Palmer amaranth is well known to have higher impact on cotton yields than late established plants [@macrae2013]. In the northern states, Palmer amaranth impact on the agroecossystem is recent. Studies investigating Palmer amaranth in those locations is limited due to the plant classification as noxious weed [@yu2021]. Nonetheless, the continuous Palmer amaranth dispersal and potential establishment across northern United States is concerning and warrant investigations on species morphology in such environments. Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, biomass production, and height of Palmer amaranth growing under corn, soybean and fallow at two timings across five locations in the United States Midwest.

@@

Palmer amaranth height and biomass were performed with a linear mixed model using *lmer* function from "lme4" package [@bates2015]. Plant height and biomass were transformed to meet model assumption of normality. In the model, agroecosystem (crop, soybean, fallow) was the fixed effect and year nested with location the random effects. Analysis of variance was performed with *anova* function from "car" package [@fox2018]. Marginal means and compact letter display were estimated with *emmeans* and *cld* from packages "emmeans" and multcomp [@hothorn2008].

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern compared to fallow and soybean at first cohort (Figure 2A). The 10% cumulative Palmer amaranth flowering in soybean, fallow and corn occurred at the end of June. Palmer amaranth reached 10% flowering in soybean at doy 180, which was slightly different from fallow (doy 180.9; *P* = 0.01) and corn (doy 181.7; *P* = 0.00). The 50% Palmer amaranth cumulative flowering occurred in July. For example, Palmer amaranth reached 50% in fallow at doy 193.4, followed by soybean (doy 194.8), corn (doy 206.6). Similar trend was observed at 90% Palmer amaranth cumulative flowering. Palmer amaranth growing in corn reached 90% flowering at doy 252.6 (early September), which was 37.8 and 32.2 days after Palmer amaranth 90% flowering in fallow and soybean, respectively.

@@

Palmer amaranth cumulative flowering at second cohort ranged from mid July to mid September (Figure 2). All cumulative flowering comparisons among agroecossystems were significant different (*P* < 0.00) Palmer amaranth growing in fallow resulted in earlier flowering time compared to soybean and corn. Palmer amaranth growing in fallow reached 10%, 50%, and 90% flowering time at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth growing in soybean reached 10% flowering at doy 210.9, which was 6 days prior to corn (*P*-value = 0.00). Similar trend was observed at 50% flowering, whereas Palmer amaranth reached 50% flowering in corn (doy 233.0) 4 days after soybeans (doy 228.9; *P* = 0.00). The 90% Palmer amaranth cumulative flowering occurred at same day in corn (260.9) and soybean (260.5; *P* = 0.66).

@@

Our study showed that Palmer amaranth flowering time, biomass and height varied within agroecossystems and cohort timing. In general, Palmer amaranth produced more biomass and taller plants when growing at first cohort rather than second cohort. At first cohort, resources (e.g., soil nutrients) and conditions (e.g., light) were more timely available for both species, crop and weed. High biomass and taller Palmer amaranth plants is likely a weed strategy to compete for light in between crop rows in absence of canopy. In such condition, Palmer amaranth showed an extraordinary plasticity to adapt upon the agroecosystem. This is evident when comparing Palmer amaranth shape, and its extended flowering pattern when growing into corn compared to soybean. In corn, Palmer amaranth was taller and thinner when growing in corn. The Palmer amaranth competition strategy was likely to mimic the crop grow and development (Figure 4). These results are concerning as breeding more competitive crop varieties is likely to select more competitive Palmer amaranth biotypes [@bravo2017].

Palmer amaranth grow and development at second cohort was limited due to the crop competitive ability at advanced development stages. Palmer amaranth was transplanted when corn canopy was nearly closed, which reduced Palmer amaranth competitiveness. As a result, Palmer amaranth height and biomass was lower compared to its first cohort. Under crop canopy, Palmer amaranth flowering pattern was near to similar into corn and soybean. Palmer amaranth growing without crop competition produced highest amounts of biomass and lowest extended flowering pattern. The Palmer amaranth strategy in bareground was to invest biomass in growing plant width and height. Nonetheless, Palmer amaranth produced 31% less biomass in second cohort compared to first cohort timing. In a bareground study, early emerged Palmer amaranth without competition was 50% taller than late emerged plants [@webster2015]. These results suggests that crop competition is not the only factor limiting late Palmer amaranth establishment. The limited growth of Palmer amaranth at second cohort is likely a reduced plant response to thermal units (e.g, growing degree days). It is hypothesize that reduced day length contributed to smaller plants at second cohort as well as shorter flowering period. A study in North Carolina and Illinois predicted that less than 10% Palmer amaranth seedlings emergence occurred after June [@piskackova2021]. In addition, Palmer amaranth negative impact on soybean [@korres2020] and cotton [@webster2015] yields was higher when plants were established near to crop planting.

@@

>>>>

In the southeastern US, early growing Palmer amaranth is well known to have higher impact on cotton yields than late established plants [@macrae2013]. In the northern states, Palmer amaranth impact on the agroecossystem is recent. Studies investigating Palmer amaranth in those locations is limited due to the plant classification as noxious weed species [@yu2021]. Nonetheless, the continuous Palmer amaranth dispersal and potential establishment across northern United States is concerning and warrant investigations on species morphology in such environments. Understanding Palmer amaranth biology and growing strategies under different agroecossystems can enhance our knowledge on species adaptation. It can also aid on designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and design resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, biomass production, and height of Palmer amaranth growing under corn, soybean and fallow at two timings across five locations in the United States Midwest.

@@

Palmer amaranth height and biomass were performed with a linear mixed model using *lmer* function from "lme4" package [@bates2015]. Plant height and biomass were transformed to meet model assumption of normality. In the model, agroecosystem (crop, soybean, fallow) was the fixed effect and year nested with location the random effects. Analysis of variance was performed with *anova* function from "car" package [@fox2018]. Marginal means and compact letter display were estimated with *emmeans* and *cld* from packages "emmeans" and "multcomp" [@hothorn2008].

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern compared to fallow and soybean at first cohort (Figure 2A). Nonetheless, the 10% cumulative Palmer amaranth flowering in soybean, fallow and corn occurred at the end of June. Palmer amaranth reached 10% flowering in soybean at doy 180, which was slightly different from fallow (doy 180.9; *P* = 0.01) and corn (doy 181.7; *P* = 0.00). The 50% Palmer amaranth cumulative flowering occurred in July. For example, Palmer amaranth reached 50% flowering in fallow at doy 193.4, followed by soybean (doy 194.8), corn (doy 206.6). Similar trend was observed at 90% Palmer amaranth cumulative flowering. Palmer amaranth growing in corn reached 90% flowering at doy 252.6 (early September), which was 37.8 and 32.2 days after Palmer amaranth 90% flowering in fallow and soybean, respectively.

@@

Palmer amaranth cumulative flowering at second cohort ranged from mid July to mid September (Figure 2B). Palmer amaranth growing in fallow resulted in earlier flowering time compared to soybean and corn. Palmer amaranth growing in fallow reached 10%, 50%, and 90% flowering time at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth growing in soybean reached 10% flowering at doy 210.9, which was 6 days prior to corn (*P*-value = 0.00). Similar trend was observed at 50% flowering, whereas Palmer amaranth reached 50% flowering in corn (doy 233.0) 4 days after soybeans (doy 228.9; *P* = 0.00). The 90% Palmer amaranth cumulative flowering occurred at same day in corn (260.9) and soybean (260.5; *P* = 0.66).

@@

Our study showed that Palmer amaranth flowering time, biomass and height varied within agroecossystems and cohort timing. In general, Palmer amaranth produced more biomass and taller plants when growing at first cohort rather than second cohort. At first cohort, resources (e.g., soil nutrients) and conditions (e.g., light) were more timely available for both species, crop and weed. High biomass and taller Palmer amaranth plants are likely a weed strategy to compete for light in between crop rows in absence of canopy. In such condition, Palmer amaranth showed an extraordinary plasticity to adapt upon the agroecosystem. This is evident when comparing Palmer amaranth shape, and its extended flowering pattern when growing into corn compared to soybean. Palmer amaranth was taller and thinner when growing within corn. The Palmer amaranth competition strategy was likely to mimic the crop grow and development (Figure 4). These results are concerning as breeding more competitive crop varieties is likely to select more competitive Palmer amaranth biotypes [@bravo2017].

Palmer amaranth grow and development at second cohort was limited due to the crop competitive ability at advanced development stages. Palmer amaranth was transplanted when corn canopy was nearly closed, which reduced Palmer amaranth competitiveness. As a result, Palmer amaranth height and biomass was lower compared to its first cohort. Under crop canopy, Palmer amaranth flowering pattern was near to similar in corn and soybean. Palmer amaranth growing without crop competition produced highest amounts of biomass and less extended flowering pattern. The Palmer amaranth strategy in bareground was to invest biomass in growing plant width and height. Nonetheless, Palmer amaranth produced 31% less biomass in second cohort compared to first cohort timing. In a bareground study, early emerged Palmer amaranth without competition was 50% taller than late emerged plants [@webster2015]. These results suggests that crop competition is not the only factor limiting late Palmer amaranth establishment. The limited growth of Palmer amaranth at second cohort is likely a reduced plant response to thermal units (e.g, growing degree days). It is hypothesize that reduced day length contributed to smaller plants at second cohort as well as shorter flowering period. A study in North Carolina and Illinois predicted that less than 10% Palmer amaranth seedlings emergence occurred after June [@piskackova2021]. In addition, Palmer amaranth negative impact on soybean [@korres2020] and cotton [@webster2015] yields was higher when plants were established near to crop planting.

@@

Palmer amaranth flowering pattern was variable

### 2021-08-05 Manuscript/Manuscript.tex
JH

In the southeastern US, early growing Palmer amaranth is well known to
have higher impact on cotton yields than late established plants (MacRae
et al., 2013). In the northern states, Palmer amaranth impact on the
agroecossystem is recent. Studies investigating Palmer amaranth in those
locations is limited due to the plant classification as noxious weed (Yu
et al., 2021). Nonetheless, the continuous Palmer amaranth dispersal and
potential establishment across northern United States is concerning and
warrant investigations on species morphology in such environments.
Understanding Palmer amaranth biology and growing strategies under
different agroecossystems can enhance our knowledge on species
adaptation. It can also aid on designing proactive and ecological
tactics to limit the species range expansion, reduce its negative
impact, and design resilient and sustainable farming systems (MacLaren
et al., 2020). Therefore, the objective of this study was to investigate
the flowering pattern, biomass production, and height of Palmer amaranth
growing under corn, soybean and fallow at two timings across five
locations in the United States Midwest.

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern
compared to fallow and soybean at first cohort (Figure 2A). The 10\%
cumulative Palmer amaranth flowering in soybean, fallow and corn
occurred at the end of June. Palmer amaranth reached 10\% flowering in
soybean at doy 180, which was slightly different from fallow (doy 180.9;
\emph{P} = 0.01) and corn (doy 181.7; \emph{P} = 0.00). The 50\% Palmer
amaranth cumulative flowering occurred in July. For example, Palmer
amaranth reached 50\% in fallow at doy 193.4, followed by soybean (doy
194.8), corn (doy 206.6). Similar trend was observed at 90\% Palmer
amaranth cumulative flowering. Palmer amaranth growing in corn reached
90\% flowering at doy 252.6 (early September), which was 37.8 and 32.2
days after Palmer amaranth 90\% flowering in fallow and soybean,
respectively.

Palmer amaranth cumulative flowering at second cohort ranged from mid
July to mid September (Figure 2). All cumulative flowering comparisons
among agroecossystems were significant different (\emph{P} \textless{}
0.00) Palmer amaranth growing in fallow resulted in earlier flowering
time compared to soybean and corn. Palmer amaranth growing in fallow
reached 10\%, 50\%, and 90\% flowering time at day 203.8, 214.4, and
232.2, respectively. Palmer amaranth growing in soybean reached 10\%
flowering at doy 210.9, which was 6 days prior to corn (\emph{P}-value =
0.00). Similar trend was observed at 50\% flowering, whereas Palmer
amaranth reached 50\% flowering in corn (doy 233.0) 4 days after
soybeans (doy 228.9; \emph{P} = 0.00). The 90\% Palmer amaranth
cumulative flowering occurred at same day in corn (260.9) and soybean
(260.5; \emph{P} = 0.66).

@@

Our study showed that Palmer amaranth flowering time, biomass and height
varied within agroecossystems and cohort timing. In general, Palmer
amaranth produced more biomass and taller plants when growing at first
cohort rather than second cohort. At first cohort, resources (e.g., soil
nutrients) and conditions (e.g., light) were more timely available for
both species, crop and weed. High biomass and taller Palmer amaranth
plants is likely a weed strategy to compete for light in between crop
rows in absence of canopy. In such condition, Palmer amaranth showed an
extraordinary plasticity to adapt upon the agroecosystem. This is
evident when comparing Palmer amaranth shape, and its extended flowering
pattern when growing into corn compared to soybean. In corn, Palmer
amaranth was taller and thinner when growing in corn. The Palmer
amaranth competition strategy was likely to mimic the crop grow and
development (Figure 4). These results are concerning as breeding more
competitive crop varieties is likely to select more competitive Palmer
amaranth biotypes (Bravo et al., 2017).

Palmer amaranth grow and development at second cohort was limited due to
the crop competitive ability at advanced development stages. Palmer
amaranth was transplanted when corn canopy was nearly closed, which
reduced Palmer amaranth competitiveness. As a result, Palmer amaranth
height and biomass was lower compared to its first cohort. Under crop
canopy, Palmer amaranth flowering pattern was near to similar into corn
and soybean. Palmer amaranth growing without crop competition produced
highest amounts of biomass and lowest extended flowering pattern. The
Palmer amaranth strategy in bareground was to invest biomass in growing
plant width and height. Nonetheless, Palmer amaranth produced 31\% less
biomass in second cohort compared to first cohort timing. In a
bareground study, early emerged Palmer amaranth without competition was
50\% taller than late emerged plants (Webster and Grey, 2015). These
results suggests that crop competition is not the only factor limiting
late Palmer amaranth establishment. The limited growth of Palmer
amaranth at second cohort is likely a reduced plant response to thermal
units (e.g, growing degree days). It is hypothesize that reduced day
length contributed to smaller plants at second cohort as well as shorter
flowering period. A study in North Carolina and Illinois predicted that
less than 10\% Palmer amaranth seedlings emergence occurred after June
(Piskackova et al., 2021). In addition, Palmer amaranth negative impact
on soybean (Korres et al., 2020) and cotton (Webster and Grey, 2015)
yields was higher when plants were established near to crop planting.

>>>>

In the southeastern US, early growing Palmer amaranth is well known to
have higher impact on cotton yields than late established plants (MacRae
et al., 2013). In the northern states, Palmer amaranth impact on the
agroecossystem is recent. Studies investigating Palmer amaranth in those
locations is limited due to the plant classification as noxious weed
species (Yu et al., 2021). Nonetheless, the continuous Palmer amaranth
dispersal and potential establishment across northern United States is
concerning and warrant investigations on species morphology in such
environments. Understanding Palmer amaranth biology and growing
strategies under different agroecossystems can enhance our knowledge on
species adaptation. It can also aid on designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and design resilient and sustainable farming systems
(MacLaren et al., 2020). Therefore, the objective of this study was to
investigate the flowering pattern, biomass production, and height of
Palmer amaranth growing under corn, soybean and fallow at two timings
across five locations in the United States Midwest.

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern
compared to fallow and soybean at first cohort (Figure 2A). Nonetheless,
the 10\% cumulative Palmer amaranth flowering in soybean, fallow and
corn occurred at the end of June. Palmer amaranth reached 10\% flowering
in soybean at doy 180, which was slightly different from fallow (doy
180.9; \emph{P} = 0.01) and corn (doy 181.7; \emph{P} = 0.00). The 50\%
Palmer amaranth cumulative flowering occurred in July. For example,
Palmer amaranth reached 50\% flowering in fallow at doy 193.4, followed
by soybean (doy 194.8), corn (doy 206.6). Similar trend was observed at
90\% Palmer amaranth cumulative flowering. Palmer amaranth growing in
corn reached 90\% flowering at doy 252.6 (early September), which was
37.8 and 32.2 days after Palmer amaranth 90\% flowering in fallow and
soybean, respectively.

Palmer amaranth cumulative flowering at second cohort ranged from mid
July to mid September (Figure 2B). Palmer amaranth growing in fallow
resulted in earlier flowering time compared to soybean and corn. Palmer
amaranth growing in fallow reached 10\%, 50\%, and 90\% flowering time
at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth growing in
soybean reached 10\% flowering at doy 210.9, which was 6 days prior to
corn (\emph{P}-value = 0.00). Similar trend was observed at 50\%
flowering, whereas Palmer amaranth reached 50\% flowering in corn (doy
233.0) 4 days after soybeans (doy 228.9; \emph{P} = 0.00). The 90\%
Palmer amaranth cumulative flowering occurred at same day in corn
(260.9) and soybean (260.5; \emph{P} = 0.66).

@@

Our study showed that Palmer amaranth flowering time, biomass and height
varied within agroecossystems and cohort timing. In general, Palmer
amaranth produced more biomass and taller plants when growing at first
cohort rather than second cohort. At first cohort, resources (e.g., soil
nutrients) and conditions (e.g., light) were more timely available for
both species, crop and weed. High biomass and taller Palmer amaranth
plants are likely a weed strategy to compete for light in between crop
rows in absence of canopy. In such condition, Palmer amaranth showed an
extraordinary plasticity to adapt upon the agroecosystem. This is
evident when comparing Palmer amaranth shape, and its extended flowering
pattern when growing into corn compared to soybean. Palmer amaranth was
taller and thinner when growing within corn. The Palmer amaranth
competition strategy was likely to mimic the crop grow and development
(Figure 4). These results are concerning as breeding more competitive
crop varieties is likely to select more competitive Palmer amaranth
biotypes (Bravo et al., 2017).

Palmer amaranth grow and development at second cohort was limited due to
the crop competitive ability at advanced development stages. Palmer
amaranth was transplanted when corn canopy was nearly closed, which
reduced Palmer amaranth competitiveness. As a result, Palmer amaranth
height and biomass was lower compared to its first cohort. Under crop
canopy, Palmer amaranth flowering pattern was near to similar in corn
and soybean. Palmer amaranth growing without crop competition produced
highest amounts of biomass and less extended flowering pattern. The
Palmer amaranth strategy in bareground was to invest biomass in growing
plant width and height. Nonetheless, Palmer amaranth produced 31\% less
biomass in second cohort compared to first cohort timing. In a
bareground study, early emerged Palmer amaranth without competition was
50\% taller than late emerged plants (Webster and Grey, 2015). These
results suggests that crop competition is not the only factor limiting
late Palmer amaranth establishment. The limited growth of Palmer
amaranth at second cohort is likely a reduced plant response to thermal
units (e.g, growing degree days). It is hypothesize that reduced day
length contributed to smaller plants at second cohort as well as shorter
flowering period. A study in North Carolina and Illinois predicted that
less than 10\% Palmer amaranth seedlings emergence occurred after June
(Piskackova et al., 2021). In addition, Palmer amaranth negative impact
on soybean (Korres et al., 2020) and cotton (Webster and Grey, 2015)
yields was higher when plants were established near to crop planting.
